A Switching Planner for Combined Task and Observation Planning

### 2011-05-03 code/subarchitectures/planner.sa/trunk/src/latex/AAAI11/switching.tex
small changes

We now describe our {\em switching} planning system that operates
according to the continual planning paradigm. The system {\em
switches} in the sense that planning and plan execution proceed in
interleaved sessions in which the {\em base planner} is either {\em
sequential} or {\em decision-theoretic}.  The first session is
sequential, and begins when a DTPDDL description of the current
problem and domain are posted to the system.  During a sequential
session a serial plan is computed that corresponds to one
execution-{\em trace} in the underlying decision-process. That trace
is a reward-giving sequence of process actions and {\em assumptive}
actions. Each {\em assumptive} action corresponds to an assertion
about the truth of a fact that is unknown at plan time -- e.g. that a
box of cornflakes is located on the corner bench in the kitchen. The
trace specifies a plan and characterises a {\em deterministic
approximation} (see~\cite{yoon:etal:2008}) of the underlying process
in which that plan is valuable. Traces are computed by a
cost-optimising classical planner which trades off action costs, goal
rewards, and determinacy. Execution of a trace proceeds according to
the process actions in the order that they appear in the trace. If,
according to the underlying belief-state, the outcome of the next
action scheduled for execution is not predetermined above a threshold
(here 95\%), then the system switches to a DT session.

Because online DT planning is impractical for the size of problem we
are interested in, DT sessions plan in a small abstract problem
defined in terms of the trace from the proceeding sequential
session. This abstract state-space is characterised by a limited
number of propositions, chosen due to their {\em relevance} to the
action whose scheduled execution triggered the switch to the DT
session.  To allow the DT planner to judge assumptions from the trace,
we add {\em disconfirm} and {\em confirm} actions to the problem for
each of them. Those yield a relatively small reward/penalty if the
corresponding judgement is true/false. If a judgement action is
scheduled for execution, then the DT session is terminated, and a new
sequential session begins.

@@

As we only consider deterministic-action POMDPs, all state
uncertainty is expressed in the $(\pp{:init})$ declaration. This
declaration is used by our approach to define the starting state for
sequential sessions, and the set of state-assumptions available to
sequential planning.  Writing \#\ if the value of a proposition is
unspecified, taking the $(\pp{:init})$ example from the previous
section, we have the following assumptions:

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.7 & kitchen & \# &  kitchen\\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

\noindent Each assumption corresponds to one distinct {\em
relaxed} visitation of the root term. Here, a conjunctive term is
visited iff its atomic subterms are visited, and zero or one of its
immediate probabilistic subterms are visited. For a sequential
session, the starting state corresponds to an assumption (usually
unique) with probability $1$.\footnote{Without a loss of generality,
we assume conjunctive terms list atoms that are true with
probability 1 according to a visitation, and that actions do not
have negative preconditions.}  In our example, that starting state is:

\small
\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{R2D2})~\pp{kitchen}),\\
\;\;(=(\pp{is-in}~\pp{box})~\#), (=(\pp{is-in}~\pp{cup})~\#)\}.
\end{array}
\]
\normalsize

To represent state assumptions we augment the problem posed during
a sequential session with an \emph{assumptive action} $\assumptiveS{i}$ for
each element $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
We also add constraints that forbid scheduling of
assumptions about facts after actions with preconditions or effects
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Let $\prob_i$ be the probability that the $i^{th}$ sequenced
action, $\action_i$, from a trace of state-action pairs
$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
does not transition to $\state^\bot$, and
$\reward(\state_i, \action_i)$ be the instantaneous reward received
for executing action $\action_i$ in state $\state_i$. The optimal
trace has value:

@@

\noindent where $\$(T_i)$ (resp. $\hat\$(T_i)$) is a
small positive (negative) numeric quantity which captures
the utility the agent receives for correctly (incorrectly) rejecting
an assumption.
%%
In terms of action physics, a disconfirm action can only be executed
once, and otherwise is modelled as a self-transformation.
%%
We only consider {\em relevant} assumptions when constructing the
abstract model.  If \switchAction\ is the action that switched the
system to a DT session, then an assumption $\assumptiveS{i}$ is {\em
relevant} if it is necessary for the outcome of \switchAction\ to be
determined.  Continuing our simplified example, if the trace is:

\noindent Taking the switching action \switchAction\ to be
$(\pp{look}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
relevant for this session, and therefore we exclude the corresponding
disconfirm action from the abstract decision process.

Given \switchAction, we also include another once-only self-transition
action $\actions.\poss(\switchAction)$, a \emph{confirmation action}
with the reward property:

@@

%%%%

In order to guarantee fast DT sessions, those plan in an abstract
process determined by the current trace and underlying belief-state.
%%
The abstract process posed to the DT planner is constructed by first
constraining as statically false all propositions except those which
are true with probability 1, or which are the subject of {\em
relevant} assumptions. For example, taking the above trace and
switching action {\texttt (look~box~office)}, the underlying
belief in Fig.~\ref{fig:beliefs}B would determine a fully
constrained belief given by Fig.~\ref{fig:beliefs}C.
%%
Next, static constraints are removed, one proposition at a time, until
the number of states that can be true with non-zero probability in the
initial belief of the abstract process reaches a given threshold.
%%
In detail, for each statically-false proposition we compute the {\em
entropy} of the state assumptions of the current trace {\em
conditional} on that proposition. 
%%
Let $X$ be a set of propositions and $2^X$ the powerset of $X$, then taking

@@

A low $\entropy(X|y)$ value suggests that knowing the truth value of
$y$ is useful for determining whether or not some assumptions $X$
hold. When removing a static constraint on propositions during the
abstract process construction, $y_i$ is considered before $y_j$ if
$\entropy(X|y_i)~<~\entropy(X|y_j)$. For example, if the serial plan
assumes the robot is in a kitchen, then propositions about the
contents of kitchens, e.g. {\texttt (=(is-in~cup)kitchen)}, are
added to characterise the abstract process' states. Taking
assumption $X$ to be {\texttt (=(is-in~box)kitchen)}, in relaxing
static constraints the following entropies are calculated:

@@

\noindent Therefore, the first static constraint to be relaxed is for
\texttt{(=(is-in~milk)office)}, or
equivalently \texttt{(=(is-in~milk)kitchen)}, giving a refined
abstract belief state depicted in
Fig.~\ref{fig:beliefs}A. Summarising, if for Fig.\ref{fig:beliefs}B
the DT session is restricted to belief-states with fewer than $8$
elements, then the starting belief-state of the DT session does not
mention a ``cup''.

\begin{figure}
\tiny
\begin{tabularx}{\columnwidth}{XX}
\hline
\vspace{1ex}
\scriptsize (A) Partially constrained abstract \mbox{\hspace{4ex}belief-state} & 
\vspace{1ex}\scriptsize \hspace{-4ex}(B) Underlying DTPDDL belief \\
\vspace{-2ex} 
\begin{tabtt}
(\=:init (=(is-in~R2D2)kitchen) \+ \\
       (\=.6(and\=(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)~kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office)))
\end{tabtt} &
\vspace{-2ex} 
\begin{tabtt}
\hspace{-6ex}(\=:init (=(is-in~R2D2)office) \+ \\
       (\=.6(and\=(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office))) \-\-\-\\
       (\=.6(=(is-in~cup)office) \+ \\
      .4(=(is-in~cup)kitchen))) \\
\end{tabtt} \\
\vspace{-6ex}\scriptsize(C) Fully constrained abstract \mbox{\hspace{5ex}belief-state} & \\
\vspace{-5ex} \begin{tabtt}
(\=:init (=(is-in~R2D2)kitchen) \+ \\
       (\=.6(=(is-in~box)kitchen))) \\
\end{tabtt} & \\
\hline
\end{tabularx}
\caption{Simplified examples of abstract belief-states from
       DT sessions. \label{fig:beliefs}}
\normalsize
\end{figure}

>>>>

We now describe our {\em switching} planning system that operates
according to the continual planning paradigm. The system {\em
  switches} in the sense that planning and plan execution proceed in
interleaved sessions in which the {\em base planner} is either {\em
  sequential} or {\em decision-theoretic}.  The first session is
sequential, and begins when a DTPDDL description of the current
problem and domain are posted to the system.  During a sequential
session a serial plan is computed that corresponds to one
execution-{\em trace} in the underlying decision-process. That trace
is a reward-giving sequence of process actions and {\em assumptive}
actions. Each {\em assumptive} action corresponds to an assertion
about some facts that are unknown at plan time -- e.g. that a box of
cornflakes is located on the corner bench in the kitchen. The trace
specifies a plan and characterises a {\em deterministic approximation}
(see~\cite{yoon:etal:2008}) of the underlying process in which that
plan is valuable. Traces are computed by a cost-optimising classical
planner which trades off action costs, goal rewards, and
determinacy. Execution of a trace proceeds according to the process
actions in the order that they appear in the trace. If, according to
the underlying belief-state, the outcome of the next action scheduled
for execution is not predetermined above a threshold (here 95\%), then
the system switches to a DT session.

%% due to their {\em relevance} to the
%% action whose scheduled execution triggered the switch to the DT
%% session.

Because online DT planning is impractical for the size of problem we
are interested in, DT sessions plan in a small abstract problem
defined in terms of the trace from the proceeding sequential session.
This abstract state-space is characterised by a limited number of
propositions, chosen because they relate evidence about assumptions
in the trace.  To allow the DT planner to judge
assumptions from the trace, we add {\em disconfirm} and {\em confirm}
actions to the problem for each of them. Those yield a relatively
small reward/penalty if the corresponding judgement is true/false. If
a judgement action is scheduled for execution, then the DT session is
terminated, and a new sequential session begins.

@@

%% Writing \#\ if the value of a proposition is unspecified,
%% taking the $(\pp{:init})$ example from the previous section, we have
%% the following assumptions:

%% \small
%% \begin{tabular}{cccc}
%% \hline
%% Probability & (is-in Robot)  & (is-in box)  & (is-in cup) \\
%% \hline
%% .7 & kitchen & \# &  kitchen\\
%% .3 & kitchen & \# & office \\
%% .8 & kitchen & office & \# \\
%% .2 & kitchen & kitchen & \# \\
%% 1.0 & kitchen & \# & \# \\
%% \hline
%% \end{tabular}
%% \normalsize

%% \noindent Each assumption corresponds to one distinct {\em
%% relaxed} visitation of the root term. Here, a conjunctive term is
%% visited iff its atomic subterms are visited, and zero or one of its
%% immediate probabilistic subterms are visited.

As we only consider deterministic-action POMDPs, all state uncertainty
is expressed in the $(\pp{:init})$ declaration. This declaration is
used by our approach to define the starting state for sequential
sessions, and the set of assumptive actions available to sequential
planning. Without a loss of generality we also suppose that actions do not have negative preconditions. For a sequential
session the starting state corresponds to the set of facts that are
true with probability $1$. Continuing our example, that starting state
is the singleton:

\small
\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{Robot})\pp{kitchen})\}.
\end{array}
\]
\normalsize

To represent state assumptions we augment the problem posed during a
sequential session with an \emph{assumptive action} $\assumptiveS{i}$
for each element, $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.  We also add
constraints that forbid scheduling of assumptions about facts after
actions with preconditions or effects that mention those facts. For
example, the robot cannot assume it is plugged into a power source
immediately after it unplugs itself.  Executing $\assumptiveS{i}$ in a
state $\state$ effects a transition to a successor state
$\state^{T_i}$, the union of $\state$ with atomic terms from $T_i$.
For example, consider the following sequential plan:

\noindent Applying the first action in $\state_0$ yields:

\small
\[
\hspace{-1ex}\begin{array}{l}
\{(=(\pp{is-in}~\pp{Robot})\pp{kitchen}),(=(\pp{is-in}~\pp{box})\pp{kitchen})\}
\end{array}
\]
\normalsize

\noindent with a probability of $0.8$. The assumed state before the
scheduled execution of action $(\pp{look}~\pp{box}~\pp{kitchen})$ is:

\small
\[
\hspace{-1ex}\begin{array}{l}
\{(=(\pp{is-in}~\pp{Robot})\pp{kitchen}),
(=(\pp{is-in}~\pp{box})\pp{kitchen}), \\\;\;(=(\pp{is-in}~\pp{cup})\pp{office})\}
\end{array}
 \]
\normalsize

and has a probability of $0.24$.

To describe the optimisation criteria used during sequential sessions
we model $\assumptiveS{i}$ probabilistically, supposing that its
application in state $\state$ effects a transition to $\state^{T_i}$
with probability $\prob_i$, and to $\state^\bot$ with probability $1 -
\prob_i$. State $\state^\bot$ is an added sink. Taking $\prob_i$ to be
the probability that the $i^{th}$ sequenced action, $\action_i$, from
a trace of state-action pairs $\langle \state_0, \action_0,\state_1,
\action_1,.., \state_N \rangle$ does not transition to $\state^\bot$,
then the optimal sequential plan has value:

@@

\noindent where $\$(T_i)$ (resp. $\hat\$(T_i)$) is a small positive
(negative) numeric quantity which captures the utility the agent
receives for correctly (incorrectly) rejecting an assumption.  In
terms of action physics, a disconfirm action can only be executed
once, and otherwise is modelled as a self-transformation.  We only
consider {\em relevant} assumptions when constructing the abstract
model.  If \switchAction\ is the action that switched the system to a
DT session, then an assumption $\assumptiveS{i}$ is {\em relevant} if
it is necessary for the outcome of \switchAction\ to be
determined. For example, taking the switching action \switchAction\ to
be $(\pp{look}~\pp{box}~\pp{kitchen})$ from our earlier sequential
plan example, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{office}))$ is not
relevant, and therefore we exclude the corresponding disconfirm action
from the abstract decision process. Given \switchAction, we also
include another once-only self-transition action
$\actions.\poss(\switchAction)$, a \emph{confirmation action} with the
reward property:

@@

In order to guarantee fast DT sessions, those plan in an abstract
process determined by the current trace and underlying belief-state.
The abstract process posed to the DT planner is constructed by first
constraining as statically false all propositions except those which
are true with probability 1, or which are the subject of {\em
  relevant} assumptions. For example, taking the above trace and
switching action {\texttt \mbox{(look~box~office)}}, the underlying
belief in Fig.~\ref{fig:beliefs}B would determine a fully constrained
belief given by Fig.~\ref{fig:beliefs}A.  Next, static constraints are
removed, one proposition at a time, until the number of states that
can be true with non-zero probability in the initial belief of the
abstract process reaches a given threshold.  In detail, for each
statically-false proposition we compute the {\em entropy} of the
relevant assumptions of the current trace {\em conditional} on that
proposition.  Let $X$ be a set of propositions and $2^X$ the powerset
of $X$, then taking

@@

A low $\entropy(X|y)$ value suggests that knowing the truth value of
$y$ is useful for determining whether or not some assumptions $X$
hold. When removing a static constraint on propositions during the
abstract process construction, $y_i$ is considered before $y_j$ if
$\entropy(X|y_i)~<~\entropy(X|y_j)$. For example, if the serial plan
assumes the box is in the kitchen, then propositions about the
contents of kitchens containing a box,
e.g. $(=(\pp{is-in}~\pp{milk})\pp{kitchen})$, are added to
characterise the abstract process' states. Taking a relevant
assumption $X$ to be $(=(\pp{is-in}~\pp{box})\pp{kitchen})$, in
relaxing static constraints the following entropies are calculated:

@@

\noindent Therefore, the first static constraint to be relaxed is for
\texttt{(=(is-in~milk)office)}, or
equivalently \texttt{(=(is-in~milk)kitchen)}, giving a refined
abstract belief state depicted in
Fig.~\ref{fig:beliefs}C. Summarising, if for Fig.\ref{fig:beliefs}B
the DT session is restricted to belief-states with fewer than $8$
elements, then the starting belief-state of the DT session does not
mention a ``cup''.

\begin{figure}
\tiny
\begin{tabularx}{\columnwidth}{XX}
\hline
\scriptsize\vspace{0.5ex}(A) Fully constrained belief & 
\vspace{0.5ex}\hspace{-4.0ex}(C) Partially constrained belief\\
\vspace{-3.0ex}\begin{tabtt}
(\=:init (=(is-in~Robot)kitchen) \+ \\
       (\=.6(=(is-in~box)kitchen))) \\
\end{tabtt} & 
\vspace{-3.0ex}\begin{tabtt}
\hspace{-5.0ex}(\=:init (=(is-in~Robot)kitchen) \+ \\
       (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)~kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office)))
\end{tabtt}\\
\vspace{-16.5ex}(B) Underlying DTPDDL belief & \\ 
\vspace{-17.0ex}\begin{tabtt}
(\=:init (=(is-in~Robot)kitchen) \+ \\
       (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office))) \-\-\-\\
       (\=.6(=(is-in~cup)office) \+ \\
      .4(=(is-in~cup)kitchen))) \\
\end{tabtt} & \\
\hline
\end{tabularx}
\caption{Simplified examples of belief-states from
       DT sessions. \label{fig:beliefs}}
\normalsize
\end{figure}

%% \begin{figure}
%% \tiny
%% \begin{tabularx}{\columnwidth}{XX}
%% \hline
%% \vspace{1ex}
%% \scriptsize (A) Partially constrained belief & %abstract belief-state & 
%% \vspace{1ex}\scriptsize \hspace{-4ex}(B) Underlying DTPDDL belief \\
%% \vspace{-2ex} 
%% \begin{tabtt}\hspace{-4ex}
%% (\=:init (=(is-in~Robot)kitchen) \+ \\
%%        (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
%%         (\=.9(=(is-in~milk)~kitchen))\+\\
%%          .1(=(is-in~milk)office))\-\-  \\
%%        .4(and(=(is-in~box)office) \+\\
%%         (.1(=(is-in~milk)kitchen))\+\\
%%         .9(=(is-in~milk)office)))
%% \end{tabtt} &
%% \vspace{-2ex} 
%% \begin{tabtt}
%% \hspace{-4ex}(\=:init (=(is-in~Robot)office) \+ \\
%%        (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
%%         (\=.9(=(is-in~milk)kitchen))\+\\
%%          .1(=(is-in~milk)office))\-\-  \\
%%        .4(and(=(is-in~box)office) \+\\
%%         (.1(=(is-in~milk)kitchen))\+\\
%%         .9(=(is-in~milk)office))) \-\-\-\\
%%        (\=.6(=(is-in~cup)office) \+ \\
%%       .4(=(is-in~cup)kitchen))) \\
%% \end{tabtt} \\
%% \vspace{-8ex}\scriptsize(C) Fully constrained belief & \\
%% \vspace{-7.5ex} \begin{tabtt}
%% (\=:init (=(is-in~Robot)kitchen) \+ \\
%%        (\=.6(=(is-in~box)kitchen))) \\
%% \end{tabtt} & \\
%% \hline
%% \end{tabularx}
%% \caption{Simplified examples of abstract belief-states from
%%        DT sessions. \label{fig:beliefs}}
%% \normalsize
%% \end{figure}
